---
title: "Time Series: Practical Issues"
subtitle: "FPP3, Chapter 13"
author: "Robert W. Walker"
institute: "AGSM"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina=3, fig.width=8, fig.height=5, warning=FALSE, message=FALSE, comment='', cache=TRUE)
library(tidyverse)
library(fpp3)
library(purrr)
library(gganimate)
library(tsibble)
options(scipen=4)
library(tidyquant)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringan)
library(xaringanthemer)
style_mono_accent(base_color = "#43418A", text_font_google   = google_font("Fira Sans", "300", "300i"))
```

---
# Practical Issues in Forecasting

## Models for different frequencies

---
### Models for annual data

 * ETS, ARIMA, Dynamic regression

---
### Models for quarterly data

 * ETS, ARIMA/SARIMA, Dynamic regression, Dynamic harmonic regression, STL+ETS, STL+ARIMA

---
### Models for monthly data

 * ETS, ARIMA/SARIMA, Dynamic regression, Dynamic harmonic regression, STL+ETS, STL+ARIMA

---
### Models for weekly data

 * ARIMA/SARIMA, Dynamic regression, Dynamic harmonic regression, STL+ETS, STL+ARIMA, TBATS

---
### Models for daily, hourly and other sub-daily data

 * ARIMA/SARIMA, Dynamic regression, Dynamic harmonic regression, STL+ETS, STL+ARIMA, TBATS

---
# Ensuring forecasts stay within limits

---
## Positive forecasts: Use logs

```{r, fig.height=3.8}
library(fpp3)
eggs <- as_tsibble(fma::eggs)
eggs %>%
  model(ETS(log(value) ~ error("A") + trend("A") + season("N"))) %>%
  forecast(h=50) %>%
  autoplot(eggs)
```

---
# Combining Forecasts
## Forecast combinations
### Clemen (1989)

"The results have been virtually unanimous: combining multiple forecasts leads to increased forecast accuracy. $\dots$ In many cases one can make dramatic performance improvements by simply averaging the forecasts."

---
## Forecast combinations

```{r}
aus_cafe <- aus_retail %>%
  filter(Industry == "Cafes, restaurants and catering services") %>%
  summarise(Turnover = sum(Turnover))
fc <- aus_cafe %>%
  filter(Month <= yearmonth("2013 Sep")) %>%
  model(
    ETS = ETS(Turnover),
    ARIMA = ARIMA(Turnover)
  ) %>%
  mutate(
    Combination = (ETS + ARIMA)/2
  ) %>%
  forecast(h = "5 years")
```


---
## Forecast combinations

```{r combineplot, echo=TRUE, fig.height=4.2}
fc %>% autoplot(aus_cafe, level = NULL) +
  labs(x = "Year", y = "$ billion",
       title = "Australian monthly expenditure on eating out")
```

---
## Forecast combinations

```{r combineaccuracy, dependson="combine1"}
fc %>% accuracy(aus_cafe)
```

---
# Missing Values

**Functions which can handle missing values**

 * `ARIMA()`
 * `TSLM()`
 * `NNETAR()`
 * `VAR()`
 * `FASSTER()`

**Models which cannot handle missing values**

 * `ETS()`
 * `STL()`
 * `TBATS()`
 
---
### What to do?

 1. Model section of data after last missing value.
 2. Estimate missing values with `interpolate()`.
 
---
## Missing values

```{r}
gold <- as_tsibble(forecast::gold)
gold %>% autoplot(value)
```

---
## Missing values

```{r, fig.height=3}
gold_complete <- gold %>%
  model(ARIMA(value)) %>%
  interpolate(gold)
gold_complete %>%
  autoplot(value, colour = "red") +
  autolayer(gold, value)
```

---
## Outliers

```{r, fig.height=3.4}
fit <- gold %>%
  model(ARIMA(value))
augment(fit) %>%
  mutate(stdres = .resid/sd(.resid, na.rm=TRUE)) %>%
  filter(abs(stdres) > 10)
```

